Crop diversification in a game against nature in a mediterranean agro-ecosystem

### 2020-01-25 risk-percption.Rmd
more plots of subjective probability

```{r, eval = FALSE}
readxl::read_xls('Data/41467_2014_BFncomms4953_MOESM580_ESM.xls', range = 'A2:O370') %>%
  select(`Site Location`, Taxon, `ky cal BP`, `GY (t*ha-1)`) %>%
  filter(!is.na(`GY (t*ha-1)`)) %>% 
  group_by(`Site Location`, Taxon) %>%
  summarise(yield_mean = mean(`GY (t*ha-1)`)) %>%
  spread(Taxon, yield_mean)
```

@@

@@

```{r}
prior <- function(m,n){
a = n * m
b = n * (1 - m)
dom <- seq(0,1,0.005)
val <- dbeta(dom,a,b)
return(data.frame('x'=dom, 'y'=val))
}

tibble(year = 1:20,
       risk = .5) %>%
  mutate(belief = map2(risk, year, ~prior(.x, .y))) %>%
  unnest(cols = c(belief)) %>%
  ggplot(aes(x, y)) +
  geom_line() +
  facet_wrap( ~year) +
  theme_bw() +
  ggtitle('Subjective beliefs about drought risk', 'Beta prior, varying sample size')

#prior(.5, ) %>% plot
```

```{r}
likelihood <- function(N,Y){
a <- Y + 1
b <- N - Y + 1
dom <- seq(0,1,0.005)
val <- dbeta(dom,a,b)
return(data.frame('x'=dom, 'y'=val))
}

#likelihood(1, 1)%>% plot
```

```{r}
posterior <- function(m,n,N,Y){
a <- Y + (n*m) -1
b <- N - Y + (n*(1-m)) - 1
dom <- seq(0,1,0.005)
val <- dbeta(dom,a,b)
return(data.frame('x'=dom, 'y'=val))
}

#dbeta(seq(0,1,0.005), -1, 0)
#prior(.5, 2) %>% plot
#posterior(.5, 3, 1, 1) %>% plot
```

@@

```{r}
learn <- function(agents, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_normal; barley_yield <- barley_normal

  agents %>%
    # farmer's turn to decide what to plant
      mutate(yield_exp = expected_yield(risk),
         barley_exp = expected_yield_barley(risk),
         wheat_exp = expected_yield_wheat(risk),
         expected_yield = barley_exp + wheat_exp,
         crop = case_when(barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat',
                              barley_exp == wheat_exp ~ 'either'),
         prediction = map2(alpha, beta, ~rbeta(100, .x, .y)), # sample 100 as if agent is thinking in percentages
         wheat_prop = map_dbl(prediction, ~sum(. < indiff$risk)) / 100) %>%
      select(-prediction) %>%
  # nature's turn to decide realized yields
    mutate(payoff = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop),
           alpha = drought + alpha,
           beta = beta + 1 - drought,
           mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}
```

Run an experiment. Simulate a time series with varying drought frequency.
```{r}
droughts <- c((runif(25) < .1),(runif(25) < .5), runif(50) < .2)
```

```{r}
simulation1 <- accumulate(droughts, ~learn(.x, .y), .init = agents) %>% 
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time != 1)
```

```{r, echo = TRUE}
ggplot(simulation1, aes(time, risk)) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()
```

@@

```{r}
reproduce <- function(individuals) {
  individuals %>%
    left_join(life_table, by = 'age') %>% 
    mutate(fertility_reduction = pgamma(payoff, shape = fertility_shape, scale = fertility_scale),
           reproduced = rbernoulli(n(), fertility_rate / 2 * fertility_reduction)) %>% # divide by two to make everyone female
    filter(reproduced == TRUE) %>% # if nrows == 0, will give a (for some reason unsuppressible) warning
    tally %>% # count births per household
    uncount(n) %>% # repeat rows based on birth count per household
    mutate(age = 0, alpha = 1.5, beta = 1.5, risk = alpha / (alpha + beta), crop = 'barley') %>%
    bind_rows(individuals, .) %>%
    ungroup # check for filled na's here?
}

die <- function(individuals) {
  individuals %>%
        left_join(life_table, by = 'age') %>% 
    mutate(survival_reduction = pgamma(pmin(1, payoff), shape = survival_shape, scale = survivor_scale),
           survived = rbernoulli(n(), survival_rate * survival_reduction)) %>%
    filter(survived == TRUE) %>%
    select(-c(fertility_rate:survived)) %>%
    mutate(age = age + 1)
}

@@

![Example model precipitation outputs from the TraCE-21k simulation at 9.5ka BP.](figures/trace_white.png)

```{r, fig.height =3, fig.width = 9, fig.cap = 'Annual risk of wheat crop failure due to drought, averaged by century.'}
sim_past %>%
  filter(between(water_year, -9500, -5500)) %>%
  mutate(time_step = ceiling(water_year / 50),
         year = time_step * 50) %>%
  group_by(year) %>%
  summarise(risk = sum(drought) / 50) %>%
  ggplot(aes(year, risk, fill = risk, color = risk)) +
  geom_hline(yintercept = indiff$risk, color = 'black', linetype = 2, width = 1.2) +
  geom_col()  +
  scale_fill_distiller(palette = 'YlOrRd', direction = 1) +
  scale_color_distiller(palette = 'YlOrRd', direction = 1)+
  theme_minimal() +
  labs(x = 'Years BP', y = 'Drought risk')  + theme(legend.position = "none")
```

```{r}
simulation4 <- accumulate(sim_past$drought, ~learn(.x, .y) %>% reproduce %>% die, .init = agents) %>% 
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time != 1)
```

So the problem now is the population dynamics are getting in the way ... and that it'd be nice to have something akin to the moran process to ensure fixed population
```{r, echo = TRUE}
ggplot(simulation4, aes(time, risk)) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()
```

```{r}
simulation4 %>%
  group_by(time) %>%
  tally %>%
  plot
```

@@

```{r}
library(ggridges)
accumulate((runif(25) < .2), ~calc_payoff(.x, drought = .y) %>% update_prop(aspiration = .6), .init = test3) %>% bind_rows(.id = 'time') %>% ggplot(aes(wheat_prop, y = as.numeric(time), group = as.numeric(time))) +
  geom_density_ridges() +
  theme_minimal()
```

```{r}
# this is for nature
calcNE <- function(risk){
  (118 - 25 * risk) / (278 - 185 * risk)
}

plot(0:10, calcNE((0:10)/10))
```

@@

# so this works as long as you're between 0 and 1
```

```{r}
yield_max_barley <- 2500
yield_max_wheat <- 3500
kcal_barley <- 3000
kcal_wheat <- 3540

@@

tibble(rainfall = (0:200) / 100,
       barley = calc_yield(rainfall, 'barley'),
       wheat = calc_yield(rainfall, 'wheat')) %>%
  gather(crop, yield, barley:wheat) %>%
  ggplot(aes(rainfall, yield, color = crop)) + 
  geom_line() +
  geom_hline(yintercept = yield_max_wheat * kcal_wheat) +
  geom_hline(yintercept = yield_max_barley * kcal_barley) +
  geom_hline(yintercept = 1016525 * 5, color ='red') + # cals for family of 5?
  geom_vline(xintercept = 0.3) +
  geom_vline(xintercept = .2)
```

>>>>

@@

```{r, fig.height =3, fig.width = 9, fig.cap = 'Annual risk of wheat crop failure due to drought, averaged by fifty-year period'}
sim_past %>%
  filter(between(water_year, -9500, -5500)) %>%
  mutate(time_step = ceiling(water_year / 50),
         year = time_step * 50) %>%
  group_by(year) %>%
  summarise(risk = sum(drought) / 50) %>%
  ggplot(aes(year, risk, fill = risk, color = risk)) +
  geom_hline(yintercept = indiff$risk, color = 'black', linetype = 2, width = 1.2) +
  geom_col()  +
  scale_fill_distiller(palette = 'YlOrRd', direction = 1) +
  scale_color_distiller(palette = 'YlOrRd', direction = 1)+
  theme_minimal() +
  labs(x = 'Years BP', y = 'Drought risk')  + theme(legend.position = "none")
```

@@

```{r}
prior <- function(m, n) {
  domain <- seq(0,1,0.005)
  val <- dbeta(domain, n * m, n * (1 - m))
  tibble('x' = domain, 'y' = val)
}

tibble(year = seq(5, 50, 5),
       risk = .5) %>%
  mutate(belief = map2(risk, year, ~prior(.x, .y))) %>%
  unnest(cols = c(belief)) %>%
  ggplot(aes(x, y)) +
  geom_line(aes(group = year, color = year), size = 1.2) +
  scale_color_viridis_c() +
  theme_bw() +
  ggtitle('Subjective beliefs about drought risk', 'Beta prior, varying sample size')
```

@@

```{r}
learn1 <- function(agents, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_normal; barley_yield <- barley_normal

  agents %>%
    # farmer's turn to decide what to plant
      mutate(yield_exp = expected_yield(risk),
         barley_exp = expected_yield_barley(risk),
         wheat_exp = expected_yield_wheat(risk),
         expected_yield = barley_exp + wheat_exp,
         crop = case_when(barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat',
                              barley_exp == wheat_exp ~ 'either'),
         prediction = map2(alpha, beta, ~rbeta(100, .x, .y)), # sample 100 as if agent is thinking in percentages
         wheat_prop = map_dbl(prediction, ~sum(. < indiff$risk)) / 100) %>%
      select(-prediction) %>%
  # nature's turn to decide realized yields
    mutate(payoff = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop),
           alpha = drought + alpha,
           beta = beta + 1 - drought,
          # mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}

learn2 <- function(agents, drought) {
  if (drought == 'TRUE') wheat_yield <- wheat_dry; barley_yield <- barley_dry
  if (drought == 'FALSE') wheat_yield <- wheat_normal; barley_yield <- barley_normal

  agents %>%
    # farmer's turn to decide what to plant
      mutate(yield_exp = expected_yield(risk),
         barley_exp = expected_yield_barley(risk),
         wheat_exp = expected_yield_wheat(risk),
         expected_yield = barley_exp + wheat_exp,
         crop = case_when(barley_exp > wheat_exp ~ 'barley',
                              barley_exp < wheat_exp ~ 'wheat',
                              barley_exp == wheat_exp ~ 'either'),
         prediction = map2(alpha, beta, ~rbeta(100, .x, .y)), # sample 100 as if agent is thinking in percentages
         wheat_prop = map_dbl(prediction, ~sum(. < indiff$risk)) / 100) %>%
      select(-prediction) %>%
  # nature's turn to decide realized yields
    mutate(payoff = wheat_yield * wheat_prop + barley_yield * (1 - wheat_prop),
           alpha = if_else(payoff < .5, 1, drought + alpha),
           beta = if_else(payoff < .5, 1, beta + 1 - drought),
          # mu = alpha / (alpha + beta),
           risk = map2_dbl(alpha, beta, ~rbeta(1, .x, .y)))
}
```

Run an experiment. Simulate a time series with varying drought frequency. This is a stepwise nonstationary process
```{r}
droughts <- c((runif(25) < .1),(runif(25) < .5), runif(50) < .2)
```

```{r}
prior2 <- function(a, b) {
  domain <- seq(0,1,0.005)
  val <- dbeta(domain, a, b)
  tibble('x' = domain, 'y' = val)
}

plot(prior2( 0, 0));plot(prior2( .5, .5));plot(prior2( 1, 1))

accumulate(droughts, 
           ~mutate(.x, alpha = .y + alpha, beta = beta + 1 - .y), 
           .init = tibble(alpha = 1.5, beta = 1.5)) %>%
  bind_rows(.id = 'time') %>%
    mutate(time = as.numeric(time), belief = map2(alpha, beta, ~prior2(.x, .y))) %>%
  unnest(cols = c(belief)) %>%
  ggplot(aes(x, y)) +
  geom_line(aes(color = time, group = time), size = .8) +
    geom_vline(xintercept = c(0.1, 0.5, 0.2), linetype = 2, color = 'red', size = .8) +
  scale_color_viridis_c() +
  theme_bw() +
  ggtitle('Subjective beliefs about drought risk', 'Beta prior, varying sample size')
```

```{r}
simulation1 <- accumulate(droughts, ~learn1(.x, .y), .init = agents) %>% 
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time != 1)

simulation2 <- accumulate(droughts, ~learn2(.x, .y), .init = agents) %>% 
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time != 1)
```

Redo these plots so its a single person, one who was born in a drought period then experiences an ameliorarion and another who experiences a shift to drought conditions later in life? basically this section neds to work more with individuals and explore that stuff

```{r, echo = TRUE}
test1 <- tibble(time = 1:100,
       risk = c(rep(0.1, 25), rep(0.5, 25), rep(0.2, 50)))

ggplot(simulation1, aes(time, risk)) +
  geom_line(data = test1) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()
```

```{r}
ggplot(simulation2, aes(time, risk)) +
  geom_line(data = test1) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()

@@

```{r}
reproduce <- function(individuals) {
  individuals %>%
    left_join(life_table, by = 'age') %>% 
    mutate(fertility_reduction = pgamma(pmin(1, payoff / 0.5), shape = fertility_shape, scale = fertility_scale),
           reproduced = rbernoulli(n(), fertility_rate / 2 * fertility_reduction)) %>% # divide by two to make everyone female
    filter(reproduced == TRUE) %>% # if nrows == 0, will give a (for some reason unsuppressible) warning
    mutate(age = 0, alpha = 1.5, beta = 1.5, risk = alpha / (alpha + beta)) %>%
    select(-c(fertility_rate:reproduced)) %>%
    bind_rows(individuals, .)
}

die <- function(individuals) {
  individuals %>%
        left_join(life_table, by = 'age') %>% 
    mutate(survival_reduction = pgamma(pmin(1, payoff / 0.5), shape = survival_shape, scale = survivor_scale),
           survived = rbernoulli(n(), survival_rate * survival_reduction)) %>%
    filter(survived == TRUE) %>%
    select(-c(fertility_rate:survived)) %>%
    mutate(age = age + 1)
}
```

@@

Another question here about initialization ... namely that we are starting with 80 year olds with no beliefs about the system?

```{r}
simulation3 <- accumulate(sim_past$drought, ~learn(.x, .y) %>% reproduce %>% die, .init = agents) %>% 
  bind_rows(.id = 'time') %>%
  mutate(time = as.numeric(time)) %>%
  filter(time != 1)
```

So the problem now is the population dynamics are getting in the way ... and that it'd be nice to have something akin to the moran process to ensure fixed population. OR recognize the importance of population dynamics and have the full continuous yields from the medland functions and just determine the chunked payoffs from there? The problem now is that if we set the food requirement to calculate the food ratio to something like 0.5, then there is no benefit to the increased yields and hence no real tradeoff between yields and risk management. perhaps just record the cumulative surplus and have that be something
```{r}
ggplot(simulation3, aes(time, risk)) +
  geom_point(alpha = .1) +
  geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()
```

```{r}
ggplot(simulation3, aes(time, payoff)) +
  geom_point(alpha = .1) +
  #geom_hline(yintercept = indiff$risk, color = 'red', linetype = 2) +
  theme_classic()
```

```{r}
simulation3 %>%
  group_by(time) %>%
  tally %>%
  plot
```

@@

@@

# so this works as long as you're between 0 and 1

```{r, eval = FALSE}
readxl::read_xls('Data/41467_2014_BFncomms4953_MOESM580_ESM.xls', range = 'A2:O370') %>%
  rename(site = `Site Location`, taxon = Taxon, age = `ky cal BP`, yield = `GY (t*ha-1)`) %>%
  select(site, taxon, age, yield) %>%
  #filter(!is.na(yield)) %>%
  mutate(yield = if_else(is.na(yield), 0, yield)) %>%
    ggplot(aes(yield)) +
  geom_histogram(aes(fill = taxon)) +
  facet_wrap(~taxon, nrow = 2)

readxl::read_xls('Data/41467_2014_BFncomms4953_MOESM580_ESM.xls', range = 'A2:O370') %>%
  rename(site = `Site Location`, taxon = Taxon, age = `ky cal BP`, yield = `GY (t*ha-1)`) %>%
  select(site, taxon, age, yield) %>%
  filter(!is.na(yield)) %>%
  filter(site == 'Akarçay Tepe') %>%
  mutate(yield = yield * 1000,
         yield = if_else(taxon == 'Hordeum vulgare', yield * 3000, yield * 3540)) %>%
group_by(taxon) %>%
  summarise(yield_mean = mean(yield), yield_sd = sd(yield)) %>%
  mutate(yield = map2(yield_mean, yield_sd, ~rnorm(100000, .x, .y))) %>%
  select(taxon, yield) %>%
  unnest(cols = c(yield)) %>% 
  mutate(yield = if_else(yield > 0, yield, 0)) %>%
  ggplot(aes(yield)) +
  geom_histogram(aes(fill = taxon)) +
  facet_wrap(~taxon, nrow = 2)
```

```{r}
rnorm(10000, 1582521,	780883.5) %>% hist
rnorm(10000, 3060657,	1461350.0) %>% hist
```

```{r}
yield_max_barley <- 2.500
yield_max_wheat <- 3.500
kcal_barley <- 1 #3000
kcal_wheat <- 3540 / 3000

@@

tibble(rainfall = (0:200) / 100,
       barley = calc_yield(rainfall, 'barley'),
       wheat = calc_yield(rainfall, 'wheat')) %>%
  gather(crop, yield, barley:wheat) %>%
  ggplot(aes(rainfall, yield, color = crop)) + 
  geom_line() +
  geom_hline(yintercept = yield_max_wheat * kcal_wheat) +
  geom_hline(yintercept = yield_max_barley * kcal_barley) +
  #geom_hline(yintercept = 1016525 * 5, color ='red') + # cals for family of 5?
  geom_vline(xintercept = 0.3)
```
